---
title: "Can you smell the difference? Identifying fungal discriminating volatiles using Random Forest classification"
author: "Lakshya Katariya"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/"
)
```

## Overview

Fungi can communicate chemically through volatile organic compounds (VOCs). In [Katariya et al. (2017)](https://link.springer.com/article/10.1007/s10886-017-0902-4),
we showed that fungus-farming termites discriminate their crop fungus
(*Termitomyces*) from a parasitic weed (*Pseudoxylaria*) using VOC profiles alone — without visual contact. 
Random Forest classification on GC-MS chemical profiles identified the candidate volatile compounds driving that discrimination.

This document has two parts:

- **Part 1** replicates the Random Forest analysis from Katariya et al. (2017) using the published supplementary data from that paper —
  *Termitomyces* vs. *Pseudoxylaria* VOC profiles.
- **Part 2** extends the same analytical framework to a broader public dataset from [Guo et al. (2021)](https://doi.org/10.1038/s42003-021-02198-8), asking whether VOC profiles discriminate fungal phyla (Ascomycota vs. Basidiomycota) across 39 species.

Random Forest is used throughout as a classification tool — the goal here is simpler — identify candidate discriminating compounds (not to build a predictive model for deployment).

---

## Setup

```{r packages}
library(tidyverse)     # dplyr, tidyr, tibble, ggplot2, forcats
library(readxl)        # read Excel files
library(varSelRF)      # iterative variable selection via Random Forest
library(janitor)       # clean_names() for column name cleaning
library(parallel)      # detect available CPU cores
library(doParallel)    # parallel backend for foreach loops
library(foreach)       # foreach loop syntax
```

---

# Part 1: Replication — *Termitomyces* vs. *Pseudoxylaria*

## Data source

VOC emission data are taken from Supplementary Table S1 of:

> Katariya L, Ramesh PB, Gopalappa T et al. (2017). *Fungus-Farming Termites Selectively Bury Weedy Fungi that Smell Different from Crop Fungi.* Journal of Chemical Ecology 43:986–995. <https://doi.org/10.1007/s10886-017-0902-4>

The table reports proportional VOC emission intensities for 8 *Termitomyces* replicates (crop fungus) and 8 *Pseudoxylaria* replicates (weedy fungus), across 41 identified volatile compounds from a representative GC-MS dataset (dataset #7 in the paper). The original analysis used 16 datasets; this replication uses dataset #7 as a representative example, consistent with the paper's Fig. 2c.

```{r load-data-p1}
# Data entered directly from Table S1 of Katariya et al. (2017)
# Rows = compounds, Columns = replicates
# T = Termitomyces (crop), P = Pseudoxylaria (weed)

voc_p1 <- tribble(
  ~compound,               ~T16a,  ~T20b,  ~T21b,  ~T77a,  ~T90b,  ~T92b,  ~TG4b,  ~TG5b,  ~P19b,  ~P20b,  ~P38b,  ~P57a,  ~P63b,  ~P65b,  ~P96a,  ~PG4a,
  "octen_3_ol",            0.105,  0.033,  0,      0,      0.070,  0.014,  0.620,  0.037,  0.019,  0.018,  0.179,  0,      0.006,  0,      0.013,  0,
  "x3_octanone",           0.181,  0.080,  0,      0,      0.243,  0.035,  0.376,  0.056,  0.011,  0.031,  0.068,  0.021,  0.025,  0.001,  0.025,  0.033,
  "x3_octanol",            0,      0,      0,      0,      0.006,  0,      0.004,  0,      0,      0,      0,      0,      0,      0,      0,      0,
  "benzaldehyde",          0,      0,      0,      0,      0.0004, 0,      0,      0,      0.008,  0,      0.009,  0.048,  0.003,  0.065,  0.009,  0.084,
  "phenylethanal",         0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0.002,  0,      0.0001, 0,      0.006,  0,
  "x2_phenylethanol",      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0.008,  0.010,  0,
  "benzene_acetonitrile",  0.173,  0,      0,      0.002,  0,      0,      0,      0.736,  0,      0,      0,      0,      0,      0,      0,      0,
  "isocyclobazzanene",     0.128,  0.077,  0,      0.372,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "bicycloelemene",        0,      0,      0.046,  0.070,  0.070,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "a_cubebene",            0.219,  0.202,  0.039,  0,      0.052,  0.039,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "cyclosativene",         0,      0,      0.037,  0,      0.029,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "b_cubebene",            0,      0.102,  0.104,  0,      0.045,  0.041,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "a_isocomene",           0,      0,      0.062,  0,      0.039,  0.029,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "aristolene",            0,      0,      0,      0,      0,      0,      0,      0,      0.133,  0.337,  0.203,  0,      0.053,  0.095,  0.092,  0.151,
  "cis_thujopsene",        0,      0,      0,      0,      0,      0,      0,      0,      0.039,  0,      0.091,  0,      0.020,  0.012,  0.031,  0,
  "b_copaene",             0,      0.122,  0.032,  0.019,  0.036,  0.047,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "b_gurjunene",           0,      0,      0,      0,      0,      0,      0,      0,      0.576,  0.065,  0,      0,      0.223,  0.384,  0.385,  0.655,
  "isobazzanene",          0,      0,      0.054,  0,      0,      0.042,  0,      0,      0,      0.016,  0,      0,      0,      0,      0,      0,
  "trans_a_bergamotene",   0,      0,      0,      0,      0,      0,      0,      0,      0.020,  0,      0,      0.517,  0.078,  0,      0.242,  0,
  "b_barbatene",           0.091,  0.063,  0.047,  0.280,  0,      0.362,  0,      0.088,  0,      0,      0,      0,      0,      0,      0,      0,
  "selina_5_11_diene",     0,      0,      0.206,  0.161,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "e_b_farnesene",         0.103,  0,      0.044,  0.046,  0.038,  0.085,  0,      0.083,  0,      0,      0,      0,      0,      0,      0,      0,
  "alloaromadendrene",     0,      0,      0,      0,      0,      0,      0,      0,      0,      0.158,  0,      0,      0.001,  0.042,  0,      0,
  "b_neoclovene",          0,      0.037,  0.015,  0,      0.017,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "b_chamigrene",          0,      0.020,  0.014,  0,      0,      0,      0,      0,      0,      0.074,  0,      0,      0,      0,      0.023,  0,
  "germacrene_d",          0,      0.093,  0.044,  0,      0.034,  0.069,  0,      0,      0,      0,      0,      0.034,  0,      0,      0,      0,
  "b_selinene",            0,      0,      0,      0,      0.015,  0,      0,      0,      0.024,  0,      0.029,  0,      0,      0.018,  0.012,  0,
  "d_selinene",            0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0.021,  0.015,  0,      0,
  "a_selinene",            0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0.054,  0,      0,      0.029,  0.038,  0,
  "epi_cubebol",           0,      0.059,  0,      0,      0.025,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "bicyclogermacrene",     0,      0,      0,      0,      0.025,  0,      0,      0,      0,      0.047,  0,      0.028,  0.025,  0,      0,      0,
  "b_curcumene",           0,      0.030,  0,      0,      0.033,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "germacrene_a",          0,      0,      0,      0.052,  0,      0.038,  0,      0,      0,      0,      0,      0,      0,      0,      0.033,  0,
  "b_bisabolene",          0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0.039,  0.057,  0.073,  0,      0,      0,
  "d_cadinene",            0,      0,      0.173,  0,      0.143,  0.185,  0,      0,      0,      0,      0,      0.025,  0.011,  0,      0,      0,
  "cadina_1_4_diene",      0,      0,      0.018,  0,      0,      0.017,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "d_cuprenene",           0,      0.081,  0.018,  0,      0.052,  0,      0,      0,      0,      0,      0,      0,      0,      0,      0,      0,
  "occidentalol",          0,      0,      0,      0,      0,      0,      0,      0,      0.012,  0.016,  0.010,  0.021,  0.007,  0.016,  0.005,  0,
  "cis_muurol_5_en_4_b_ol",0,      0,      0.045,  0,      0.026,  0,      0,      0,      0,      0.026,  0,      0,      0,      0,      0,      0,
  "viridiflorol",          0,      0,      0,      0,      0,      0,      0,      0,      0.121,  0.160,  0.122,  0.094,  0.091,  0.095,  0.019,  0.015,
  "pogostol",              0,      0,      0,      0,      0,      0,      0,      0,      0.024,  0.053,  0.196,  0.155,  0.363,  0.220,  0.058,  0.061
)

head(voc_p1)

cat("Compounds:", nrow(voc_p1), "\n")
cat("Replicates:", ncol(voc_p1) - 1, "\n")
```

## P1.1 Transpose and label

```{r transpose-p1}
df_p1 <- voc_p1 %>%
  pivot_longer(cols      = -compound,
               names_to  = "replicate",
               values_to = "emission") %>%
  mutate(fungi = as.factor(if_else(
    str_starts(replicate, "T"), "Termitomyces", "Pseudoxylaria"
  ))) %>%
  pivot_wider(names_from = compound, values_from = emission) %>%
  column_to_rownames("replicate") %>%
  mutate(across(-fungi, as.numeric))

cat("Replicates per group:\n")
print(table(df_p1$fungi))
cat("VOC features:", ncol(df_p1) - 1, "\n")
```

## P1.2 Variable selection via varSelRF

We repeat the original analysis from Katariya et al. (2017): `varSelRF` run 100 times to identify the most consistently selected minimal compound set.

```{r varselrf-p1, cache=TRUE}
n_cores <- detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

set.seed(42)

mod_p1 <- foreach(i         = 1:100,
                  .combine  = c,
                  .packages = "varSelRF") %dopar% {
                    varSelRF(xdata = df_p1[, names(df_p1) != "fungi"], Class = df_p1$fungi)$selected.model
                  }

stopCluster(cl)
registerDoSEQ()

sort(summary(as.factor(mod_p1)), decreasing = TRUE)
```

Plotting RF results identifying compounds that can explain the dissimilarities in the volatile profiles of the two fungi: 

```{r fig-varselrf-p1, fig.width=8, fig.height=4, dpi=150}
sort(summary(as.factor(mod_p1)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = fct_reorder(paste0("Combination ", row_number()), frequency)) %>%
  ggplot(aes(x = combination, y = frequency)) +
  geom_col(fill = "#2E7D6B", width = 0.5) +
  geom_text(aes(label = frequency), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(
    title    = "Top 3 most frequently selected compound sets",
    subtitle = "varSelRF repeated 100 times — Termitomyces vs. Pseudoxylaria",
    x        = NULL,
    y        = "Number of runs selected (out of 100)",
    caption  = "Data: Katariya et al. (2017) J. Chem. Ecol. 43:986–995"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9),
    axis.text.y   = element_text(size = 11)
  )
```

Compound identities for each combination:

```{r varselrf-table-p1}
sort(summary(as.factor(mod_p1)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = paste0("Combination ", row_number())) %>%
  select(combination, frequency, model)
```

## P1.3 Fit Random Forest on selected variables

`varSelRF` identifies which compounds to include but does not produce a model suitable for inference. We therefore fit a full Random Forest on the selected variables to obtain OOB error, feature importance, and the proximity matrix needed for visualisation.

```{r fit-rf-p1}
best_p1 <- names(sort(summary(as.factor(mod_p1)), decreasing = TRUE))[1]
vars_p1  <- str_split(best_p1, " \\+ ")[[1]] %>% trimws()
cat("Selected compounds:", paste(vars_p1, collapse = ", "), "\n")

set.seed(42)
rf_p1 <- randomForest(
  fungi ~ .,
  data       = df_p1 %>% select(fungi, all_of(vars_p1)),
  ntree      = 1000,
  importance = TRUE,
  proximity  = TRUE
)
print(rf_p1)
```

OOB error rate:

```{r oob-p1}
oob_p1 <- rf_p1$err.rate %>%
  as.data.frame() %>%
  slice_tail(n = 1) %>%
  pull(OOB)
cat("OOB error rate:", round(oob_p1 * 100, 1), "%\n")
```

## P1.4 Feature importance

```{r fig-importance-p1, fig.width=8, fig.height=5, dpi=150}
importance(rf_p1, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("compound") %>%
  rename(importance = MeanDecreaseAccuracy) %>%
  arrange(desc(importance)) %>%
  ggplot(aes(x = fct_reorder(compound, importance), y = importance)) +
  geom_col(fill = "#2E7D6B", width = 0.7) +
  coord_flip() +
  labs(
    title    = "VOCs discriminating Termitomyces vs. Pseudoxylaria",
    subtitle = "Random Forest: Mean Decrease in Accuracy (permutation importance)",
    x        = NULL,
    y        = "Mean decrease in accuracy",
    caption  = "Data: Katariya et al. (2017) J. Chem. Ecol. 43:986–995"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9)
  )
```

## P1.5 MDS plot

```{r fig-mds-p1, fig.width=8, fig.height=6, dpi=150}
mds_p1 <- cmdscale(1 - rf_p1$proximity, k = 2) %>%
  as.data.frame() %>%
  rename(MDS1 = V1, MDS2 = V2) %>%
  rownames_to_column("replicate") %>%
  left_join(df_p1 %>% rownames_to_column("replicate") %>% select(replicate, fungi),
            by = "replicate")

fungi_colours <- c("Termitomyces" = "#1D9E75",
                   "Pseudoxylaria" = "#D85A30")

ggplot(mds_p1, aes(
  x = MDS1,
  y = MDS2,
  colour = fungi,
  label = replicate
)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_text(size = 2.8,
            vjust = -0.8,
            alpha = 0.7) +
  scale_colour_manual(values = fungi_colours, name = "Fungus") +
  labs(
    title    = "VOC-based similarity of Termitomyces and Pseudoxylaria replicates",
    subtitle = "MDS on Random Forest proximity matrix",
    x        = "MDS axis 1",
    y = "MDS axis 2",
    caption  = "Data: Katariya et al. (2017) J. Chem. Ecol. 43:986–995"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    plot.caption    = element_text(colour = "grey50", size = 9),
    legend.position = "right"
  )
```

## P1.6 Confusion matrix

```{r confusion-p1}
rf_p1$confusion %>%
  as.data.frame() %>%
  rownames_to_column("true_group") %>%
  rename(`OOB error` = class.error)
```

## Part 1 summary

- **`r nrow(df_p1)`** replicates (8 *Termitomyces*, 8 *Pseudoxylaria*), **`r ncol(df_p1) - 1`** VOC features.
- `varSelRF` (100 runs) identifies **`r paste(vars_p1, collapse = " + ")`** as the most stable discriminating compound set — consistent with the
  sesquiterpene-driven discrimination reported in Katariya et al. (2017).
- OOB error rate: **`r round(oob_p1 * 100, 1)`%**.
- The MDS plot shows clear separation of *Termitomyces* and *Pseudoxylaria* replicates in VOC space, replicating Fig. 2c of the original paper.

---

# Part 2: Extension — Ascomycota vs. Basidiomycota (Guo et al. 2021)

Having replicated the core finding on our own data, we now ask whether the
same analytical framework reveals broader phylogenetic signal in VOC chemistry.
We apply it to 39 species across two major fungal phyla using data from
[Guo et al. (2021)](https://doi.org/10.1038/s42003-021-02198-8).

## Data source

- **Paper:** Guo et al. (2021). *Volatile organic compound patterns predict fungal trophic mode and lifestyle.* Communications Biology 4:673.
- **Data:** Supplementary Table S4 (GC-MS emission intensities) and Table S1 (phylum assignments), available at <https://osf.io/bva2q>

## P2.1 Load and prepare data

```{r load-data-p2}
# Download TableS4 from https://osf.io/bva2q and place in your working directory
data_raw <- read_excel(
  path  = "TableS4_GCMSdata.xlsx",
  # change name if required
  sheet = 1,
  skip  = 1
)

data_clean <- data_raw %>%
  select(-matches("^Mean|^SD")) %>%
  clean_names() %>%
  filter(!str_detect(compounds, "^[0-9]"))  # keep named compounds only

head(data_clean)
```

## P2.2 Extract species names

```{r species-names-p2}
species_names <- colnames(data_clean) %>%
  setdiff("compounds") %>%
  str_remove("_[0-9]{1,3}$") %>%
  unique()

species_names
```

## P2.3 Assign phylum labels

```{r phylum-labels}
phylum_labels <- tribble(
  ~species_name,      ~phylum,
  # Ascomycota
  "c_montecillanum",  "Ascomycota",
  "s_macrospora",     "Ascomycota",
  "a_alternata",      "Ascomycota",
  "a_brassicicola",   "Ascomycota",
  "a_niger",          "Ascomycota",
  "a_oryzae",         "Ascomycota",
  "b_cinerea",        "Ascomycota",
  "b_sorokiniana",    "Ascomycota",
  "c_globosum",       "Ascomycota",
  "c_herbarum",       "Ascomycota",
  "d_teres",          "Ascomycota",
  "f_culmorum",       "Ascomycota",
  "f_graminearum",    "Ascomycota",
  "f_oxysporum",      "Ascomycota",
  "g_candidum",       "Ascomycota",
  "m_bicolor",        "Ascomycota",
  "p_lilacinus",      "Ascomycota",
  "p_oxalicum",       "Ascomycota",
  "s_sclerotiorum",   "Ascomycota",
  "t_hamatum",        "Ascomycota",
  "t_reesei",         "Ascomycota",
  "t_velutinum",      "Ascomycota",
  "thz_es891",        "Ascomycota",
  "thz_ms8a1",        "Ascomycota",
  "thz_wm24a1",       "Ascomycota",
  "v_albo_atrum",     "Ascomycota",
  # Basidiomycota
  "a_porphyria",      "Basidiomycota",
  "c_glaucopus",      "Basidiomycota",
  "h_annosum",        "Basidiomycota",
  "l_bicolor",        "Basidiomycota",
  "p_chrysosporium",  "Basidiomycota",
  "p_croceum",        "Basidiomycota",
  "p_placenta",       "Basidiomycota",
  "p_squarrosa",      "Basidiomycota",
  "t_hirsuta",        "Basidiomycota",
  "t_versicolor",     "Basidiomycota",
  "t_zygodesmoides",  "Basidiomycota",
  "u_hordei",         "Basidiomycota",
  "u_nuda",           "Basidiomycota"
)

cat("Ascomycota species:", sum(phylum_labels$phylum == "Ascomycota"), "\n")
cat("Basidiomycota species:", sum(phylum_labels$phylum == "Basidiomycota"), "\n")
```

## P2.4 Filter and transpose

```{r filter-transpose-p2}
data_selected <- data_clean %>%
  select(compounds, starts_with(phylum_labels$species_name)) %>%
  filter(if_any(-compounds, ~ .x != 0))

df_p2 <- data_selected %>%
  pivot_longer(cols = -compounds,
               names_to = "species",
               values_to = "emission") %>%
  mutate(species_name = str_remove(species, "_[0-9]{1,3}$")) %>%
  left_join(phylum_labels, by = "species_name") %>%
  mutate(phylum = as.factor(phylum)) %>%
  select(-species_name) %>%
  pivot_wider(names_from = compounds, values_from = emission) %>%
  column_to_rownames("species") %>%
  mutate(across(-phylum, as.numeric)) %>%
  clean_names()

cat("Observations (replicates):", nrow(df_p2), "\n")
cat("VOC features:", ncol(df_p2) - 1, "\n")
cat("\nReplicates per phylum:\n")
print(table(df_p2$phylum))
```

## P2.5 Variable selection via varSelRF

```{r varselrf-p2, cache=TRUE}
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

set.seed(42)

mod_p2 <- foreach(i         = 1:100,
                  .combine  = c,
                  .packages = "varSelRF") %dopar% {
                    varSelRF(xdata = df_p2[, names(df_p2) != "phylum"], Class = df_p2$phylum)$selected.model
                  }

stopCluster(cl)
registerDoSEQ()

sort(summary(as.factor(mod_p2)), decreasing = TRUE)
```

Plotting RF results identifying compounds that can explain the dissimilarities in the volatile profiles of the two Phyla: 

```{r fig-varselrf-p2, fig.width=8, fig.height=4, dpi=150}
sort(summary(as.factor(mod_p2)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = fct_reorder(paste0("Combination ", row_number()), frequency)) %>%
  ggplot(aes(x = combination, y = frequency)) +
  geom_col(fill = "#2E7D6B", width = 0.5) +
  geom_text(aes(label = frequency), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(
    title    = "Top 3 most frequently selected compound sets",
    subtitle = "varSelRF repeated 100 times — Ascomycota vs. Basidiomycota",
    x        = NULL,
    y        = "Number of runs selected (out of 100)",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9),
    axis.text.y   = element_text(size = 11)
  )
```

Compound identity:

```{r varselrf-table-p2}
sort(summary(as.factor(mod_p2)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = paste0("Combination ", row_number())) %>%
  select(combination, frequency, model)
```

## P2.6 Fit Random Forest on selected variables

```{r fit-rf-p2}
best_p2  <- names(sort(summary(as.factor(mod_p2)), decreasing = TRUE))[1]
vars_p2  <- str_split(best_p2, " \\+ ")[[1]] %>% trimws()
cat("Number of selected compounds:", length(vars_p2), "\n")

set.seed(42)
rf_p2 <- randomForest(
  phylum ~ .,
  data       = df_p2 %>% select(phylum, all_of(vars_p2)),
  ntree      = 1000,
  importance = TRUE,
  proximity  = TRUE
)
print(rf_p2)
```

OOB error rate:

```{r oob-p2}
oob_p2 <- rf_p2$err.rate %>%
  as.data.frame() %>%
  slice_tail(n = 1) %>%
  pull(OOB)
cat("OOB error rate:", round(oob_p2 * 100, 1), "%\n")
```

## P2.7 Feature importance

```{r fig-importance-p2, fig.width=8, fig.height=6, dpi=150}
importance(rf_p2, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("compound") %>%
  rename(importance = MeanDecreaseAccuracy) %>%
  arrange(desc(importance)) %>%
  ggplot(aes(x = fct_reorder(compound, importance), y = importance)) +
  geom_col(fill = "#2E7D6B", width = 0.7) +
  coord_flip() +
  labs(
    title    = "VOCs discriminating Ascomycota vs. Basidiomycota",
    subtitle = "Random Forest: Mean Decrease in Accuracy (permutation importance)",
    x        = NULL,
    y        = "Mean decrease in accuracy",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9)
  )
```

## P2.8 MDS plot

```{r fig-mds-p2, fig.width=9, fig.height=7, dpi=150}
mds_p2 <- cmdscale(1 - rf_p2$proximity, k = 2) %>%
  as.data.frame() %>%
  rename(MDS1 = V1, MDS2 = V2) %>%
  rownames_to_column("species") %>%
  left_join(df_p2 %>% rownames_to_column("species") %>% select(species, phylum),
            by = "species")

phylum_colours <- c("Ascomycota" = "#D85A30",
                    "Basidiomycota" = "#1D9E75")

ggplot(mds_p2, aes(
  x = MDS1,
  y = MDS2,
  colour = phylum,
  label = species
)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_text(size = 2.6,
            vjust = -0.8,
            alpha = 0.7) +
  scale_colour_manual(values = phylum_colours, name = "Phylum") +
  labs(
    title    = "VOC-based similarity of fungal replicates by phylum",
    subtitle = "MDS on Random Forest proximity matrix",
    x        = "MDS axis 1",
    y = "MDS axis 2",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    plot.caption    = element_text(colour = "grey50", size = 9),
    legend.position = "right"
  )
```

## P2.9 Confusion matrix

```{r confusion-p2}
rf_p2$confusion %>%
  as.data.frame() %>%
  rownames_to_column("true_phylum") %>%
  rename(`OOB error` = class.error)
```

## Part 2 summary

- Phyla compared: **Ascomycota** (`r sum(df_p2$phylum == "Ascomycota")`
  replicates) vs. **Basidiomycota** (`r sum(df_p2$phylum == "Basidiomycota")` replicates).
- `varSelRF` (100 runs) identified a stable minimal discriminating compound set (Combination 1 above).
- OOB error rate: **`r round(oob_p2 * 100, 1)`%**.
- The MDS plot shows whether replicates cluster by phylum in VOC space.

---

## Overall conclusion

The Random Forest framework developed in Katariya et al. (2017) to identify "weedy scent" compounds in a termite mutualism generalises beyond the original system. In Part 1, the replication on published data confirms that sesquiterpene compounds reliably discriminate *Termitomyces* from
*Pseudoxylaria*. In Part 2, applying the same framework to 39 species across two phyla shows that VOC chemistry carries broader phylogenetic signal — suggesting that volatile profiles encode biological identity at multiple taxonomic levels.

---

## Session info

```{r session-info}
sessionInfo()
```
