---
title: "Fungal VOC profiles discriminate between phyla: a Random Forest approach"
author: "Lakshya Katariya"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/"
)
```

## Overview

Fungi communicate chemically through volatile organic compounds (VOCs). But do these scents encode something deeper — like *which phylum* a fungus belongs to?

In [Katariya et al. (2017)](https://link.springer.com/article/10.1007/s10886-017-0902-4), we showed that fungus-farming termites discriminate their crop fungus from a parasitic weed using VOC profiles alone — without visual contact. Random Forest classification on GC-MS data identified the candidate compounds driving that discrimination.

Here I apply the same analytical approach to a publicly-available dataset from [Guo et al. (2021)](https://doi.org/10.1038/s42003-021-02198-8), who measured comprehensive VOC profiles from 43 fungal species. Using phylum-level metadata from Supplementary Table S1, I compare **Ascomycota** and **Basidiomycota** — the two major fungal phyla in the dataset — and ask: do their VOC profiles differ systematically, and which compounds carry the most discriminatory signal?

Random Forest is used here as a classification tool: the goal is to identify candidate discriminating compounds, following the analytical approach in Katariya et al. (2017).

---

## Data source

- **Paper:** Guo et al. (2021). *Volatile organic compound patterns predict fungal trophic mode and lifestyle.* Communications Biology 4:673.
- **VOC data:** Supplementary Table S4 (GC-MS emission intensities),   available at <https://osf.io/bva2q>.
- **Metadata:** Supplementary Table S1 (phylum assignments), same source.

---

## Setup

```{r packages}
library(tidyverse)     # dplyr, tidyr, tibble, ggplot2, forcats
library(readxl)        # read Excel files
library(varSelRF)      # iterative variable selection via Random Forest
library(janitor)       # clean_names() for column name cleaning
library(parallel)      # detect available CPU cores
library(doParallel)    # parallel backend for foreach loops
library(foreach)       # foreach loop syntax
```

---

## 1. Load and prepare VOC data

Table S4 has compounds as rows and fungal species as columns — the transpose of what `randomForest()` expects. We drop the `Mean` and `SD` summary columns, which are not individual observations. We also retain only compounds with identified names, excluding rows where the compound is recorded only as a mass-to-charge ratio (m/z), as these cannot be meaningfully used downstream.

```{r load-data}
# Download TableS4 from https://osf.io/bva2q and place in your working directory
data_raw <- read_excel(
  path  = "TableS4_GCMSdata.xlsx",  # change name if required
  sheet = 1,
  skip  = 1        # row 1 is a description; row 2 becomes the header
)

data_clean <- data_raw %>%
  select(-matches("^Mean|^SD")) %>%   # drop summary columns
  clean_names() %>%
  filter(!str_detect(compounds, "^[0-9]"))  # keep named compounds only

head(data_clean)
```

---

## 2. Extract species names

```{r species-names}
species_names <- colnames(data_clean) %>%
  setdiff("compounds") %>%
  str_remove("_[0-9]{1,3}$") %>%
  unique()

species_names
```

---

## 3. Assign phylum labels from Supplementary Table S1

Phylum assignments are taken directly from Table S1 of Guo et al. (2021). Zygomycota species (*Mucor circinelloides*, *Mucor racemosus*, *Rhizopus oryzae*) are excluded — this analysis focuses on the Ascomycota vs. Basidiomycota comparison.

```{r phylum-labels}
phylum_labels <- tribble(
  ~ species_name,
  ~ phylum,
  # Ascomycota
  "c_montecillanum",
  "Ascomycota",
  "s_macrospora",
  "Ascomycota",
  "a_alternata",
  "Ascomycota",
  "a_brassicicola",
  "Ascomycota",
  "a_niger",
  "Ascomycota",
  "a_oryzae",
  "Ascomycota",
  "b_cinerea",
  "Ascomycota",
  "b_sorokiniana",
  "Ascomycota",
  "c_globosum",
  "Ascomycota",
  "c_herbarum",
  "Ascomycota",
  "d_teres",
  "Ascomycota",
  "f_culmorum",
  "Ascomycota",
  "f_graminearum",
  "Ascomycota",
  "f_oxysporum",
  "Ascomycota",
  "g_candidum",
  "Ascomycota",
  "m_bicolor",
  "Ascomycota",
  "p_lilacinus",
  "Ascomycota",
  "p_oxalicum",
  "Ascomycota",
  "s_sclerotiorum",
  "Ascomycota",
  "t_hamatum",
  "Ascomycota",
  "t_reesei",
  "Ascomycota",
  "t_velutinum",
  "Ascomycota",
  "thz_es891",
  "Ascomycota",
  "thz_ms8a1",
  "Ascomycota",
  "thz_wm24a1",
  "Ascomycota",
  "v_albo_atrum",
  "Ascomycota",
  # Basidiomycota
  "a_porphyria",
  "Basidiomycota",
  "c_glaucopus",
  "Basidiomycota",
  "h_annosum",
  "Basidiomycota",
  "l_bicolor",
  "Basidiomycota",
  "p_chrysosporium",
  "Basidiomycota",
  "p_croceum",
  "Basidiomycota",
  "p_placenta",
  "Basidiomycota",
  "p_squarrosa",
  "Basidiomycota",
  "t_hirsuta",
  "Basidiomycota",
  "t_versicolor",
  "Basidiomycota",
  "t_zygodesmoides",
  "Basidiomycota",
  "u_hordei",
  "Basidiomycota",
  "u_nuda",
  "Basidiomycota"
)

cat("Ascomycota species:",
    sum(phylum_labels$phylum == "Ascomycota"),
    "\n")
cat("Basidiomycota species:",
    sum(phylum_labels$phylum == "Basidiomycota"),
    "\n")
```

---

## 4. Filter and transpose to species x compounds format

Keep only columns belonging to Ascomycota and Basidiomycota. Remove compounds not detected in any retained species (all-zero rows carry no discriminatory information).

```{r filter-transpose}
selected_species <- phylum_labels$species_name

data_selected <- data_clean %>%
  select(compounds, starts_with(selected_species)) %>%
  filter(if_any(-compounds, ~ .x != 0))

# Transpose: rows = replicates, columns = compounds
df_species <- data_selected %>%
  pivot_longer(cols      = -compounds,
               names_to  = "species",
               values_to = "emission") %>%
  mutate(species_name = str_remove(species, "_[0-9]{1,3}$")) %>%
  left_join(phylum_labels, by = "species_name") %>%
  mutate(phylum = as.factor(phylum)) %>%
  select(-species_name) %>%
  pivot_wider(names_from  = compounds, values_from = emission) %>%
  column_to_rownames("species") %>%
  mutate(across(-phylum, as.numeric)) %>%
  clean_names()

cat("Observations (replicates):", nrow(df_species), "\n")
cat("VOC features:", ncol(df_species) - 1, "\n")
cat("Phyla compared:", paste(levels(df_species$phylum), collapse = " vs. "), "\n")
cat("\nReplicates per phylum:\n")
print(table(df_species$phylum))
```

---

## 5. Variable selection via varSelRF

`varSelRF` iteratively eliminates the least important variables, running a new Random Forest at each step until a minimal discriminating set is found.
We repeat this 100 times (as in Katariya et al. 2017) to identify which compound combinations are selected most consistently across runs.

This is computationally intensive. We parallelise across CPU cores, leaving one core free to avoid system instability.

```{r varselrf, cache=TRUE}
n_cores <- detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

cat("Running varSelRF across", n_cores, "cores...\n")

set.seed(42)

mod1res <- foreach(i         = 1:100,
                   .combine  = c,
                   .packages = "varSelRF") %dopar% {
                     varSelRF(xdata = df_species[, names(df_species) != "phylum"], Class = df_species$phylum)$selected.model
                   }

stopCluster(cl)
registerDoSEQ()

sort(summary(as.factor(mod1res)), decreasing = TRUE)
```

The plot below shows the three most frequently selected compound combinations across 100 runs. A dominant combination — selected in a high proportion of runs — indicates a stable, reproducible discriminating signal.

```{r fig-varselrf, fig.width=8, fig.height=4, dpi=150}
sort(summary(as.factor(mod1res)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = fct_reorder(paste0("Combination ", row_number()), frequency)) %>%
  ggplot(aes(x = combination, y = frequency)) +
  geom_col(fill = "#2E7D6B", width = 0.5) +
  geom_text(aes(label = frequency), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(
    title    = "Top 3 most frequently selected compound sets",
    subtitle = "varSelRF repeated 100 times — Ascomycota vs. Basidiomycota",
    x        = NULL,
    y        = "Number of runs selected (out of 100)",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9),
    axis.text.y   = element_text(size = 11)
  )
```

Compound identities for each combination:

```{r varselrf-table}
sort(summary(as.factor(mod1res)), decreasing = TRUE) %>%
  as.data.frame() %>%
  rownames_to_column("model") %>%
  rename(frequency = ".") %>%
  slice_head(n = 3) %>%
  mutate(combination = paste0("Combination ", row_number())) %>%
  select(combination, frequency, model)
```

---

## 6. Fit full Random Forest on selected variables

`varSelRF` identifies the minimal discriminating compound set. We extract the most frequently selected set (Combination 1) and fit a full `randomForest()` on those variables to obtain importance scores, proximity matrix, and OOB error.

```{r fit-rf}
best_model <- names(sort(summary(as.factor(mod1res)), decreasing = TRUE))[1]

selected_vars <- str_split(best_model, " \\+ ")[[1]] %>% trimws()
cat("Number of selected compounds:", length(selected_vars), "\n")

set.seed(42)

rf_model <- randomForest(
  phylum ~ .,
  data       = df_species %>% select(phylum, all_of(selected_vars)),
  ntree      = 1000,
  importance = TRUE,
  proximity  = TRUE
)

print(rf_model)
```

```{r oob-error}
oob_rate <- rf_model$err.rate %>%
  as.data.frame() %>%
  slice_tail(n = 1) %>%
  pull(OOB)

cat("Final OOB error rate:", round(oob_rate * 100, 1), "%\n")
```

---

## 7. Which VOCs discriminate Ascomycota from Basidiomycota?

Mean Decrease in Accuracy measures how much model accuracy drops when a compound's values are randomly shuffled — compounds with high scores are genuinely informative discriminators between the two phyla.

```{r feature-importance}
imp_df <- importance(rf_model, type = 1) %>%
  as.data.frame() %>%
  rownames_to_column("compound") %>%
  rename(importance = MeanDecreaseAccuracy) %>%
  arrange(desc(importance))

imp_df
```

```{r fig-importance, fig.width=8, fig.height=6, dpi=150}
ggplot(imp_df, aes(x = fct_reorder(compound, importance), y = importance)) +
  geom_col(fill = "#2E7D6B", width = 0.7) +
  coord_flip() +
  labs(
    title    = "VOCs discriminating Ascomycota vs. Basidiomycota",
    subtitle = "Random Forest: Mean Decrease in Accuracy (permutation importance)",
    x        = NULL,
    y        = "Mean decrease in accuracy",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    plot.caption  = element_text(colour = "grey50", size = 9)
  )
```

---

## 8. VOC-based replicate similarity (MDS)

The Random Forest proximity matrix captures which replicates co-occur in terminal nodes most often — a non-parametric measure of similarity in VOC space. Multidimensional scaling (MDS) projects this into two dimensions, coloured by phylum. Clear separation of phyla in MDS space would indicate that VOC profiles reliably encode phylogenetic identity — consistent with the principle demonstrated in Katariya et al. (2017).

```{r fig-mds, fig.width=9, fig.height=7, dpi=150}
mds_coords <- cmdscale(1 - rf_model$proximity, k = 2)

mds_df <- mds_coords %>%
  as.data.frame() %>%
  rename(MDS1 = V1, MDS2 = V2) %>%
  rownames_to_column("species") %>%
  left_join(df_species %>%
              rownames_to_column("species") %>%
              select(species, phylum),
            by = "species")

phylum_colours <- c("Ascomycota" = "#D85A30",
                    "Basidiomycota" = "#1D9E75")

ggplot(mds_df, aes(
  x = MDS1,
  y = MDS2,
  colour = phylum,
  label  = species
)) +
  geom_point(size = 3.5, alpha = 0.85) +
  geom_text(size = 2.6,
            vjust = -0.8,
            alpha = 0.7) +
  scale_colour_manual(values = phylum_colours, name = "Phylum") +
  labs(
    title    = "VOC-based similarity of fungal replicates by phylum",
    subtitle = "MDS on Random Forest proximity matrix",
    x        = "MDS axis 1",
    y        = "MDS axis 2",
    caption  = "Data: Guo et al. (2021) Commun. Biol. 4:673"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(colour = "grey40", size = 10),
    plot.caption    = element_text(colour = "grey50", size = 9),
    legend.position = "right"
  )
```

---

## 9. Robustness check: confusion matrix

The confusion matrix reveals whether misclassification is symmetric across phyla or concentrated in one direction — important context for interpreting the OOB error rate.

```{r confusion-matrix}
rf_model$confusion %>%
  as.data.frame() %>%
  rownames_to_column("true_phylum") %>%
  rename(`OOB error` = class.error)
```

---

## Summary

- Phyla compared: **Ascomycota** (`r sum(df_species$phylum == "Ascomycota")` replicates) vs. **Basidiomycota** (`r sum(df_species$phylum == "Basidiomycota")` replicates).
- `varSelRF` repeated 100 times identified a stable minimal set of   discriminating VOCs (Combination 1 in the plot above).
- A Random Forest model fit on the selected compounds achieved an OOB error rate of **`r round(oob_rate * 100, 1)`%**.
- The MDS proximity plot shows whether replicates cluster by phylum in VOC
  space — consistent with the principle demonstrated in Katariya et al.
  (2017): that biological identity is encoded in volatile chemistry.

---

## Session info

```{r session-info}
sessionInfo()
```
